---
title: "NESMR Plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nesmr_plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#͜#>",
  fig.width = 7,
  fig.height = 7
)
```


## Simulation Example
```{r}
set.seed(13)
library(esmr)
library(GWASBrewer)
G <- matrix(
  c(0, 0, 0, 0,
    0.25, 0, 0, 0,
    0, 0, 0,  0,
    0, -0.15, 0.2, 0),
  nrow = 4,
  byrow = 4
)

d <- ncol(G)

B_full <- matrix(1, d, d)
B_lower <- lower.tri(G) + 0
B_correct <- (G != 0) + 0

h2 <- 0.2
J <- 5000
N <- 20000
pi_J <- 0.1
alpha <- 5e-8

dat <- sim_mv(
    G = G,
    N = N,
    J = J,
    h2 = h2,
    pi = pi_J,
    sporadic_pleiotropy = TRUE,
    est_s = TRUE
)
```

## Discovery

```{r}
suppressWarnings(capture.output(mh_results <- mh_graph_explore(dat), file = nullfile()))

kableExtra::kable(discovery_summary(mh_results), digits = 4, row.names = F)
```

### Discovery Plots - ELBO

```{r, fig.height = 6}
library(ggplot2)

plot(mh_results, plot_type = "norm_elbo")
plot(mh_results, plot_type = "cum_norm_elbo")
plot(mh_results, plot_type = "both")
```

## Individual Graph Plots

Individual graph objects extend [tidygraphs](https://tidygraph.data-imaginist.com/) which provide an easy interface that works with [ggraph](https://ggraph.data-imaginist.com/index.html) for plotting.

We can get the best graph from the discovery results.

```{r}
best_graph <- top_i_graph(mh_results)

print(best_graph)
```

### Adjacency Graph

Plot trinary graph (+1,0,-1) for positive and negative effects

```{r}
plot_layered_topo(best_graph)
```

### Weighted Graph

```{r}
plot_layered_topo(best_graph, plot_type = "beta", weight = "direct_effect")
```

### Weighted Graph - Tile

```{r}
plot_nesmr_tile(best_graph)
```

### Edge Inclusion Probabilities - Tile

```{r}
edge_inclusion_tile_plot(mh_results)
```

## Graph Differences

### Graph Differences - Beta difference

```{r}
best_graph <- top_i_graph(mh_results)
next_graph <- top_i_graph(mh_results, i = 2)

plot_graph_differences(best_graph, next_graph, diff_type = "beta")
```

```{r}
# Adjacency (edge presence) differences
plot_graph_differences(best_graph, next_graph, diff_type = "adj")
```