---
title: "NESMR Plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NESMR Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#͜#>",
  fig.width = 7,
  fig.height = 7
)
```

# Introduction

In this vignette, we demonstrate how to use the NESMR package for plotting and visualizing network structures inferred from GWAS summary statistics.
We provide a simulation example, perform network discovery, and illustrate various plotting functions available in the package.


## Simulation Example

We simulate GWAS summary statistics for traits that have a specified network structure using [GWASBrewer](https://github.com/jean997/GWASBrewer).

```{r}
set.seed(13)
library(esmr)
library(GWASBrewer)
G <- matrix(
  c(0, 0, 0, 0,
    0.25, 0, 0, 0,
    0, 0, 0,  0,
    0, -0.15, 0.2, 0),
  nrow = 4,
  byrow = 4
)

d <- ncol(G)

B_full <- matrix(1, d, d)
B_lower <- lower.tri(G) + 0
B_correct <- (G != 0) + 0

h2 <- 0.2
J <- 5000
N <- 20000
pi_J <- 0.1
alpha <- 5e-8

dat <- sim_mv(
    G = G,
    N = N,
    J = J,
    h2 = h2,
    pi = pi_J,
    sporadic_pleiotropy = TRUE,
    est_s = TRUE
)
```

## Discovery

```{r}
suppressWarnings(capture.output(mh_results <- mh_graph_explore(dat), file = nullfile()))

kableExtra::kable(
  head(discovery_summary(mh_results), n = 10),
  digits = 4)
```

### Discovery Plots - ELBO

```{r norm-elbo-plots, fig.height = 6}
library(ggplot2)

plot(mh_results, plot_type = "norm_elbo")
plot(mh_results, plot_type = "cum_norm_elbo")
plot(mh_results, plot_type = "both")
```

## Individual Graph Plots

Individual graph objects extend [tidygraphs](https://tidygraph.data-imaginist.com/) which provide an easy interface that works with [ggraph](https://ggraph.data-imaginist.com/index.html) for plotting.

We can get the best graph from the discovery results.

```{r best-graph}
best_graph <- top_i_graph(mh_results)

print(best_graph)
```

### Adjacency Graph

Plot trinary graph (+1,0,-1) for positive and negative effects

```{r best-graph-adj}
plot_layered_topo(best_graph)
```

### Weighted Graph

```{r best-graph-beta}
plot_layered_topo(best_graph, plot_type = "beta", weight = "direct_effect")
```

### Weighted Graph - Tile

The default plot for the weighted adjacency matrix plots the the tiles in matrix ordering. This can be controlled with the `x_axis_position` argument which accepts `top` or `bottom`.
The variables are topologically ordered by default in the same order as the `plot_layered_topo` function.

```{r tile-best-graph-beta}
plot_nesmr_tile(best_graph, x_axis_position = "top", node_order = 4:1)
```

```{r, tile-best-graph-beta-bottom}
plot_nesmr_tile(best_graph, x_axis_position = "bottom", node_order = 4:1)
```

### Edge Inclusion Probabilities - Tile

Edge inclusion probabilities are the approximate posterior inclusion probability, or the sum of all of the normalized ELBOs in which that edge is present in a graph.

```{r eip-tile-plot}
edge_inclusion_tile_plot(mh_results, node_order = 4:1)
```

## Graph Differences

We can compare the different graphs returned from NESMR discovery using a few different plotting function.

### Graph Differences - Beta difference

```{r graph-diff-plot, fig.width = 10, fig.height = 5}
library(gridExtra)
best_graph <- top_i_graph(mh_results)
next_graph <- top_i_graph(mh_results, i = 2)

coords <- layered_topo_sort_coords(best_graph)
p1 <- plot_nesmr_graph(best_graph, coords = coords, plot_type = "beta") + ggtitle("Best Graph")
p2 <- plot_nesmr_graph(next_graph, coords = coords, plot_type = "beta") + ggtitle("Next Graph")
grid.arrange(p1, p2, nrow = 1)
```

We can use the `-` operation on graphs to compute the differences between the graphs.
This returns a tidygraph object which has edge comparisons for each edge in both graphs.
These are columns in the tidygraph edges:

  - `adj_diff` Adjacency difference. For a single graph, an edge present is considered a 1 and missing is considered zero. Then `adj_diff` is subtraction from these binary variables resulting in possible values of -1, 0, 1 representing edge in second graph only, edge present or absent in both graphs, or edge present in first graph only
  - `sign_diff` Differences in the signs of the effects
    - `-2 = "Sign Flip: - → +"`
    - `-1 = "Edge Removed"`
    - `0 = "No Change"`
    - `1 = "Edge Added"`
    - `2 = "Sign Flip: + → -"`
  - `direct_effect` is now the difference of effects

```{r graph-diff}
print(best_graph - next_graph)
```

These differences can be plotted in matrix/tile plotting functions

```{r graph-diff-plot-beta}
plot_graph_differences(best_graph, next_graph, diff_type = "beta", node_order = 4:1)
```

```{r graph-diff-plot-adj}
# Adjacency (edge presence) differences
plot_graph_differences(best_graph, next_graph, diff_type = "adj", node_order = 4:1)
```