---
title: "nesmr_intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nesmr_intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(esmr)
```


## Linear Structural Equations Models and Graphs

A linear structural equation model describes the relationships between a set of variables using linear equations. The relationships can be represented graphically as directed acyclic graphs (DAGs), where nodes represent variables and edges represent linear relationships.



## Simulating Data

We simulate GWAS summary statistics for traits that have a specified network structure using GWASBrewer (TODO: Link).


```{r sim_data}
library(GWASBrewer)
library(esmr)

G <- matrix(
  c(0, 0, 0, 0,
    0.25, 0, 0, 0,
    0, 0, 0,  0,
    0, -0.15, 0.2, 0),
  nrow = 4,
  byrow = 4
)

d <- ncol(G)

# Trait heritability
h2 <- 0.2
# Total number of SNPs
J <- 5000
# Number of samples in each GWAS
N <- 20000
# Proportion of SNPs with causal effects on each trait
pi_J <- 0.1
# Statistical significance threshold
alpha <- 5e-8

dat <- sim_mv(
    G = G,
    N = N,
    J = J,
    h2 = h2,
    pi = pi_J,
    sporadic_pleiotropy = TRUE,
    est_s = TRUE
)
```

## Variant selection

We select variants that are significantly associated with at least one trait.
Since we have simulated data, we can use the true Z-scores to select variants to avoid winner's curse bias.
We use a significance threshold of $\alpha = 5\times 10^{-8}$ to select variants. The `minp` function computes the minimum p-value across traits for each variant, and we select variants with a minimum p-value below the threshold.
This means that we include any variant that has at least one trait with a p-value below $5\times 10^{-8}$.


```{r select_variants}
Ztrue <- with(dat, beta_marg/se_beta_hat)
pval_true <- 2*pnorm(-abs(Ztrue))
minp <- apply(pval_true, 1, min)
ix <- which(minp < alpha)
```

## Fitting the model

NESMR fits a linear structural equation model to the selected variants.
We control what this model looks like by specifying the graph structure via an adjacency matrix.
An entry $(i,j)$ in the adjacency matrix `B` specifies that we want to estimate a causal effect of trait $i$ on trait $j$.
We assume for now that `B` needs to be a directed acyclic graph (DAG), i.e. that there are no cycles in the graph.
That is, we need to be able to order the traits in a way so that `B` is lower triangular.

Under the NESMR assumptions, if we specify a template `B` that has the true causal relationships as a subgraph, then the model will be identifiable and we can estimate the causal effects.

